Genetic Dilution erodes productivity: Exploring farmers' low adoption levels of improved maize in Ethiopia

### 2023-03-23 paper.tex
Updates from Overleaf
mar 23 changes

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in  developing new  maize varieties, evidence shows that few farmers consistently adopt this new technology over time and that average effects mask considerable hetergeneity in impacts. Misclassification of seed is a crucial problem in these contexts, as non-classical measurement error can bias the impacts. We take a step further and study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find peculiar results driven by bias due to the misclassification of seed. We exploit unique information on maize DNA-fingerprinting collected over the same areas in 2018. We find that misclassification in self-reported hybrid seed adoption mask heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data. Positive returns to adoption are found for those farmers that use higher-purity germplasm, drought-tolerant maize, and newly released varieties. When we account for misclassification in our heterogeneity analysis, we find that the puzzle is resolved. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium. The implications of our findings speak to the need for policies to better target context and geography, expand accessibility of improved seeds, and make higher yields varieties more inclusive.

@@

Low adoption of hybrid seed, despite its benefits, has been a longstanding question in the study of agricultural technologies. Several answers have been proposed to explain such low adoption of improved varieties. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. An alternative mechanism for a lack of improvement has been the misclassification of seed and seed dilution (\citealt{wossen2022misperceiving}; \citealt{kosmowski2019varietal}). Net returns to improved seed is often lower than expected in the long term because a large percentage of farmers use diluted improved seed varieties, or ones that have been recycled over many generations (\citealt{euler2022because}; \citealt{wossen2019estimating}). Evidence suggests that yield improvement and disease resistance wanes over time \citep{Abate2015-rj}. Conversely, given the ubiquity of improved seed varieties over the course of decades of access in Ethiopia, some farmers may use improved seed without knowing it, leading to inferior yields than conventional seed due to a lack of the application of complementary inputs.

Misclassification of improved seed and its implications for measuring productivity impacts has been a relatively new branch of study, with the advent of DNA fingerprinting data. For instance, \cite{wossen2019estimating} use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria. In the context of hybrid seed adoption, however, there is evidence to suggest that productivity impacts of seed affect households heterogeneously \citep{Suri2011-oi}, and that average impacts to adoption may not be as informative. The observed effect of adoption may be low because households that would have a comparative advantage in adoption of improved seed are constrained by a lack of access to improved seed or infrastructure. Additionally, the non-classical measurement error introduced by misclassification can bias heterogeneity in ways that may attenuate or even reverse productivity impacts. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of  improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties. Specifically, extension services and on-farm field trials, seed variety characteristics and rainfall have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}. Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}. Other important determinants of heterogeneous effects include risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx}. We are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed.

@@

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \input{results/tables/summary}
    \vspace*{-4em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.7\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. Maize farmers only.
            \end{tablenotes}
        \end{tabular}
    \end{table}  
    \newpage
}}

@@

\begin{figure}[htpb]
    \centering
    \caption{Adoption rates of improved varieties in 2018 (by self-report, purity level, crop type, source and year of release)}\label{fig:adoption_r4}
    \includegraphics[width=.7\textwidth]{results/figures/adoption_r4.png}
    \begin{table}[H]
    \centering
        \begin{tabular}{p{0.7\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Shares represent the proportion of adopters among maize farmers. Acronyms include: SR1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; SR2  [=1 if New, =0 if 2nd gen, recycled or traditional]; DNA70 - Improved if purity $>=$70$\%$; DNA90 - Improved if purity $>=$90$\%$; DNA95 - Improved if purity $>=$95$\%$; HYB - Improved DNA, [=1 Hybrid, 0=Open pollinated]; DTM - Drought tolerant Maize [=1 DTM, 0=otherwise], CGIAR source [=1 if germplasm associated to a CGIAR-related thread], Exotic source [=1 if germplasm associated to an exotic source], Released 2000+ [=1 if DNA associated to a variety released on year 2000 or later], Released 2010+ [=1 if DNA associated to a variety released on year 2010 or later]. QPM is not included as there are too few households planting QPM seeds (n=6).
            \end{tablenotes}
        \end{tabular}
    \end{table}   
\end{figure}

@@

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \centering
    \input{results/tables/traj}
    \vspace*{-4em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{0.7\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. Maize farmers only. Table shows household adoption trajectories, their frequency and their share in the sample. 1 denotes adoption of improved maize seed, 0 otherwise. The first digit denotes adopting in wave 1, the second in wave 2, and the third in wave 3. 
            \end{tablenotes}
        \end{tabular}
    \end{table}  
    \newpage
}}

@@

As a last step, we introduce a misclassification-robust approach to the GRC method, which allows us to take misclassification into account when calculating comparative advantage. Using the DNA identification from ESS4 we are able to verify how often an improved seed is actually used when a farmer self-reports adopting improved seed. The methodology in this section builds on work that was done in \cite{michuda2021three}, which used confusion matrices from machine learning algorithms as a way to estimate a causal effect that was robust to misclassification. Misclassification introduces non-classical measurement error that may attenuate estimates and bias them towards 0 in the binary case. In the multi-regime case, misclassification will tend to make estimates too close to one another, leading to smaller or even wrong-signed estimates.

@@

\input{results/tables/misclass_95purity.tex}

>>>>

    \clearpage\maketitle
	\thispagestyle{empty}
	\vspace*{-2em}
    \begin{center}\begin{abstract}
    \singlespacing
			\noindent  
			Despite large investments in developing new maize varieties, evidence shows that few farmers consistently adopt them. Moreover, average positive effects on yields mask considerable heterogeneity, particularly in contexts where seed are not traceable (missclasified as new varieties). We take a step further and study misclassification in the context of heterogeneous impacts to adoption. We use a three-year panel to estimate heterogeneous returns to adoption from self-reports of hybrid seed use and find null returns from adoption. We show how this result is driven misclassification of seeds that induce bias by exploiting a unique information on maize DNA-fingerprinting. When we account for seed misclassification, we find that heterogeneity in the actual quality of the genetic material of the seeds farmers categorized as improved varieties in the self-reported data masks positive returns to adoption, with the highest returns found for farmers using higher-purity germplasm, drought-tolerant maize, and newly released varieties. Our findings point to the wide dispersion of older and genetically diluted varieties, for which poorer farmers may be paying a premium, and therefor the need for policies that expand the accessibility of improved, context and geography-specific seeds, which display higher yields.

@@

Low adoption of hybrid seed, despite its benefits, has been a longstanding question in the study of agricultural technologies. Given their capacity to provide higher and less variable yields, lack of consistent adoption has received a lot of attention from the economic literature, with several explanations developed in the literature. These include lack of liquidity and other economic constraints \citep{Carter2014-fm}, growing conditions such as the type of soil, input use, and farmer characteristics \citep{Munshi2004-og}, imperfections in credit markets \citep{Croppenstedt2003-pq}, ambiguously defined property rights \citep{Place2000-el}, lack of commitment devices \citep{Duflo2009-iv}, high transportation costs that increase the cost of agricultural inputs \citep{Byerlee2013-qk}, and improved seed supply constraints \citep{Bird2020-nt}. An alternative mechanism for a lack of improvement has been the misclassification of seed and seed genetic dilution (\citealt{wossen2022misperceiving}; \citealt{kosmowski2019varietal}). Yield improvement and disease resistance wanes over time as the genetic material is recycled \citep{Abate2015-rj}, making that the net returns from these \textit{improved} seed is lower than expected, and decreases in the long term as farmers continue to use genetically dilute seed varieties, or ones that have been recycled over many generations (\citealt{euler2022because}; \citealt{wossen2019estimating}). Moreover, as genes from improved varieties enter the regular seed market its inadvertent use can potentially lead to inferior yields since they then to require the application of complementary inputs.

% Given the ubiquity of improved seed varieties over the course of decades of access in Ethiopia,

The greater availability of DNA fingerprinting data makes possible a new revision of the role of misclassification of improved seed on the adoption of new varieties. For instance, \cite{wossen2019estimating} use DNA fingerprinting data to show that there is a 22\% bias in the productivity impact of improved cassava seed in Nigeria, suggesting that average impacts to adoption may not be as informative. Additionally, missclassification introduces a non-classical measurement error that can make that the observed productivity impact is attenuated or even reversed. This paper provides a way to account for misclassification of seed when additional, error-free data exists, such as DNA fingerprinting data. We estimate the comparative advantage of households directly using a Group Random Coefficients (GRC) Model and correct its heterogeneity estimates using misclassification rates from DNA fingerprinting data of the same plots \citep{Suri2011-oi, Tjernstrom_Emilia_Dalia_Ghanem_Oscar_Barriga_Cabanillas_Travis_J_Lybbert_Jeffrey_D_Michler_and_Aleksandr_Michuda2020-bc}.

@@

Our results contribute to the literature on the drivers of low adoption rates of improved seed varieties as we are the first to consider how heterogeneous impacts to seed adoption are affected by misclassification of seed. The consensus in this literature is that the success of improved varieties depends on a range of factors, and that the widespread adoption of improved seeds is the result of complementary conditions and inputs that allow farmers to extract the full potential of these varieties, while other factos such as risk aversion \citep{Holden2016-vy}, farm size \citep{Ghimire2015-bd}, and credit constraints \citep{Simtowe2008-jn,Balana2020-hx} also play a role.\footnote{Complementary inputs include extension services and on-farm field trials, while complementary conditions refer to seed characteristics adapted for the terrain conditions and appropriate rainfall and temperature levels. Input availability and appropriate conditions have been found to be crucial to the adoption of improved maize in Tanzania \citep{Kaliba2000-jh}}. In the specific case of Ethiopia, labor, fertilizer use, farmers’ experience with extension packages, rainfall suitability and prices have been shown to increase the opportunity cost of not adopting improved wheat varieties \citep{Wale2006-bv}.\footnote{Market access and the accessibility of extension services are also critical in the adoption of improved chickpea varieties in Ethiopia \citep{Verkaart2019-ol}}

@@

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \centering
        \input{results/tables/summary}
        \vspace*{-4em}
            \begin{table}[H]
                \centering
                \begin{tabular}{p{1\textwidth}} 
                    \begin{tablenotes}
                          \small
                          \item Note: Author's calculations using ESS. Mean and standard deviation for each variable shown. Standard deviation in parentheses. Maize farmers only. Parcel size, yield and distance variables winsorized at the 1\% level.
                    \end{tablenotes}
                \end{tabular}
        \end{table}  
\newpage
}}

@@

\begin{figure}[htpb]
    \centering
    \caption{Adoption rates of improved varieties in 2018 (by self-report, purity level, crop type, source and year of release)}\label{fig:adoption_r4}
    \includegraphics[width=.7\textwidth]{results/figures/adoption_r4.png}
    \begin{table}[H]
    \centering
        \begin{tabular}{p{1\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Shares represent the proportion of adopters among maize farmers. Acronyms include: SR1 [=1 if New, 2nd gen or recycled, =0 if Traditional]; SR2  [=1 if New, =0 if 2nd gen, recycled or traditional]; DNA70 - Improved if purity $>=$70$\%$; DNA90 - Improved if purity $>=$90$\%$; DNA95 - Improved if purity $>=$95$\%$; HYB - Improved DNA, [=1 Hybrid, 0=Open pollinated]; DTM - Drought tolerant Maize [=1 DTM, 0=otherwise], CGIAR source [=1 if germplasm associated to a CGIAR-related thread], Exotic source [=1 if germplasm associated to an exotic source], Released 2000+ [=1 if DNA associated to a variety released on year 2000 or later], Released 2010+ [=1 if DNA associated to a variety released on year 2010 or later]. QPM is not included as there are too few households planting QPM seeds (n=6).
            \end{tablenotes}
        \end{tabular}
    \end{table}   
\end{figure}

@@

\afterpage{{\footnotesize\tabcolsep=1pt  % hold it local
    \centering
    \input{results/tables/traj}
    \vspace*{-4em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{1\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Author's calculations using ESS. Maize farmers only. Table shows household adoption trajectories, their frequency and their share in the sample. 1 denotes adoption of improved maize seed, 0 otherwise. The first digit denotes adopting in wave 1, the second in wave 2, and the third in wave 3. 
            \end{tablenotes}
        \end{tabular}
    \end{table}  
    \newpage
}}

@@

As a last step, we introduce a misclassification-robust approach to the GRC method, which allows us to take misclassification into account when calculating comparative advantage. Using the DNA identification from ESS4 we are able to verify how often an improved seed is actually used when a farmer self-reports adopting improved seed. The methodology in this section builds on work that was done in \cite{michuda2021three}, which used confusion matrices from machine learning algorithms as a way to estimate a causal effect that was robust to misclassification (see Table \ref{tab:misclassification}). Misclassification introduces non-classical measurement error that may attenuate estimates and bias them towards 0 in the binary case. In the multi-regime case, misclassification will tend to make estimates too close to one another, leading to smaller or even wrong-signed estimates.

@@

\input{results/tables/misclass_95purity.tex}
\vspace*{-2em}
    \begin{table}[H]
        \centering
        \begin{tabular}{p{1\textwidth}} 
            \begin{tablenotes}
                  \small
                  \item Note: Table shows the joint distribution of misclassifying adoption of hybrid maize. Adoption is self-reported from the first three data rounds of the ESS data, while misclassification is estimated from the fourth round. Joint distribution assumes misclassification is constant and exhibits no dynamic behavior.
            \end{tablenotes}
        \end{tabular}
\end{table}

### 2023-03-23 results/tables/misclass_95purity.tex
mar 23 changes

\begin{table}
\caption{Misclassification Matrix for Seed Adoption based on DNA Fingerprinting}
\label{tab:cm}
\centering
\begin{tabular}{lllllllll}
\toprule
 & 000 & 001 & 010 & 011 & 100 & 101 & 110 & 111 \\
\midrule
000 & 0.643 & 0.102 & 0.102 & 0.016 & 0.102 & 0.016 & 0.016 & 0.003 \\
001 & 0.220 & 0.524 & 0.035 & 0.083 & 0.035 & 0.083 & 0.006 & 0.013 \\
010 & 0.220 & 0.035 & 0.524 & 0.083 & 0.035 & 0.006 & 0.083 & 0.013 \\
011 & 0.076 & 0.180 & 0.180 & 0.428 & 0.012 & 0.029 & 0.029 & 0.068 \\
100 & 0.220 & 0.035 & 0.035 & 0.006 & 0.524 & 0.083 & 0.083 & 0.013 \\
101 & 0.076 & 0.180 & 0.012 & 0.029 & 0.180 & 0.428 & 0.029 & 0.068 \\
110 & 0.076 & 0.012 & 0.180 & 0.029 & 0.180 & 0.029 & 0.428 & 0.068 \\
111 & 0.026 & 0.062 & 0.062 & 0.147 & 0.062 & 0.147 & 0.147 & 0.349 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
    \item \footnotesize Notes: Table shows results of misclassification probabilities for each trajectory in the sample, using probabilities from Table \ref{tab:a}. Misclassification probabilities calculated from DNA fingerprinting data from Wave 4 of the ESS and assumes that true adoption of hybrid seed is when DNA purity of hybrid is at least 95\%.
\end{tablenotes}
\end{table}

>>>>

\begin{table}[H]
\caption{Misclassification Matrix for Seed Adoption based on DNA Fingerprinting}
\label{tab:cm}
\centering
\begin{tabular}{lllllllll}
\toprule
 & 000 & 001 & 010 & 011 & 100 & 101 & 110 & 111 \\
\midrule
000 & 0.643 & 0.102 & 0.102 & 0.016 & 0.102 & 0.016 & 0.016 & 0.003 \\
001 & 0.220 & 0.524 & 0.035 & 0.083 & 0.035 & 0.083 & 0.006 & 0.013 \\
010 & 0.220 & 0.035 & 0.524 & 0.083 & 0.035 & 0.006 & 0.083 & 0.013 \\
011 & 0.076 & 0.180 & 0.180 & 0.428 & 0.012 & 0.029 & 0.029 & 0.068 \\
100 & 0.220 & 0.035 & 0.035 & 0.006 & 0.524 & 0.083 & 0.083 & 0.013 \\
101 & 0.076 & 0.180 & 0.012 & 0.029 & 0.180 & 0.428 & 0.029 & 0.068 \\
110 & 0.076 & 0.012 & 0.180 & 0.029 & 0.180 & 0.029 & 0.428 & 0.068 \\
111 & 0.026 & 0.062 & 0.062 & 0.147 & 0.062 & 0.147 & 0.147 & 0.349 \\
\bottomrule
\end{tabular}
\end{table}
